---
title: ClusteringMetrics
package: ClusteringMetrics
author: 
- name: Siyuan Luo
  email: roseluosy@gmail.com
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document
vignette: |
  %\VignetteIndexEntry{1_introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
library(BiocStyle)
```

# Getting started

A partition is a way of organizing a dataset into distinct, non-overlapping, and non-empty subsets. For example, a clustering is a partition. This package provides multiple approaches for comparing two partitions of the same dataset, and evaluating the alignment between a dataset’s embedding/graph representations and its partition.

Unlike a hard partition, a fuzzy partition allows data points to belong to multiple subsets with varying degrees of membership. This package further offers methods for comparing two fuzzy partitions as well as for comparing a hard partition with a fuzzy partition.

Additionally, spatial clustering can be evaluated by evaluating the hard or fuzzy partitions, or by using internal metrics. This package provides functions for all these types of analysis.

## Installation

```{r, eval=FALSE}
if (!requireNamespace("devtools", quietly = TRUE))
    install.packages("devtools")
devtools::install_github("RoseYuan/ClusteringMetrics")
```

## Example data

To showcase the main functions, we will use an example object which contains (a 
subset of) RNAseq of mouse hippocampi after Forskolin-induced long-term 
potentiation:

```{r}
suppressPackageStartupMessages({

})

```



```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```